\Title

### 2013-02-15 synmut.tex
word version for Jan

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\PCApat}{1}
\newcommand{\timedependence}{3}
\newcommand{\withinepi}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Intrapatient HIV evolution is dominated by selection on the protein level in the
arms race with the adaptive immune system. When killer T cells or antibodies
target a new epitope, the virus often escapes via nonsynonymous mutations that
impair recognition. Synonymous mutations do not affect this interplay and are
often assumed to be neutral.
We analyze longitudinal intrapatient data from the \shankaregion{} part of the
envelope gene (\env{}) and observe that synonymous derived alleles rarely
fix even though they often reach high frequencies in the viral population.
We find that synonymous mutations that disrupt base pairs in RNA stems flanking
the variable loops of gp120 are more likely to be lost than other synonymous
changes, hinting at a direct fitness effect of these stem-loop structures in the
HIV RNA.
Computational modeling indicates that these synonymous mutations have a
(Malthusian) selection coefficient of the order of $-0.002$, and that they are
brought up to high frequency by hitchhiking on neighboring beneficial
nonsynonymous alleles. The patterns of fixation of nonsynonymous mutations
suggest that escape mutations in \shankaregion{} are only transiently
beneficial, either because the immune system is catching up or because of
competition between equivalent escapes.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 2400 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous the
fixation probability and diversity observed in patients. Parameters are chosen
from uniform prior distributions as indicated by the red box (see methods).}
\label{fig:simheat}
\end{center}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, onecolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\PCApat}{1}
\newcommand{\timedependence}{3}
\newcommand{\withinepi}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Intrapatient HIV evolution is dominated by selection on the protein level in the
arms race with the adaptive immune system. When killer T cells or antibodies
target a new epitope, the virus often escapes via nonsynonymous mutations that
impair recognition. Synonymous mutations do not affect this interplay and are
often assumed to be neutral.
We analyze longitudinal intrapatient data from the \shankaregion{} part of the
envelope gene (\env{}) and observe that synonymous derived alleles rarely
fix even though they often reach high frequencies in the viral population.
We find that synonymous mutations that disrupt base pairs in RNA stems flanking
the variable loops of gp120 are more likely to be lost than other synonymous
changes, hinting at a direct fitness effect of these stem-loop structures in the
HIV RNA.
Computational modeling indicates that these synonymous mutations have a
(Malthusian) selection coefficient of the order of $-0.002$, and that they are
brought up to high frequency by hitchhiking on neighboring beneficial
nonsynonymous alleles.
Besides, the patterns of fixation of nonsynonymous mutations estimated from the
longitudinal data and comparisons with computer models
suggest that escape mutations in \shankaregion{} are only transiently
beneficial, either because the immune system is catching up or because of
competition between equivalent escapes.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.7\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=0.7\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $\pfix$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run 2400 simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous the
fixation probability and diversity observed in patients. Parameters are chosen
from uniform prior distributions as indicated by the red box (see methods).}
\label{fig:simheat}
\end{center}
\end{figure}
